---
title: "Simulating the Luria Delbrück experiment"
output:
html_document: default
pdf_document: default
---

# Introduction

This R markdown document will form the basis from which you will write your own code to simulate the fluctuation (Luria-Delbrück) experiment. After every step, we will make sure everyone is on the same page, so if you'll get stuck, don't worry too much. A big part of modelling is just having fun and playing! =)


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Setup chunk always runs when opening your RMD project. Instead of just blindly
# trying to install the packages, I first test if they are already installed by
# using `require`.

if(!require(tidyr)) install.packages("tidyr", repos = "http://cran.us.r-project.org")  
if(!require(ggplot2)) install.packages("ggplot2", repos = "http://cran.us.r-project.org") 
if(!require(patchwork)) install.packages("patchwork", repos = "http://cran.us.r-project.org")

library(dplyr)            # R data transformation package
library(ggplot2)          # Plotting
library(patchwork)        # Putting plots next to each other

theme_set(theme_bw())     # Set a theme for plotting in ggplot (but you can plot however you like ofc.)


L <- genomeLength <- 4629812              # Escherichia coli B REL606
m <- mutationRate <- 2.2*10^(-10)         # Lee et al. 2012 PNAS
P <- maxPopSize   <- 2*10^8               # max population size in the simulation


```


# Step 1) Modeling exponential growth


## Assumptions
* There is no cell death
* Every bacterium divides every time step

```{r exponential growth}

# Below is a simple function to model exponential growth. 
exponentialGrowth=function(maxPopsize){
  N <- 1 # Initial population size. In modelling, usually denoted with 'N'
  t <- 1 # Initial time point
  output <- data.frame(Time=t,Cellcount=N) # New (empty) dataframe containing two columns: Time, Cellcount
  while(N<maxPopsize){
     N <- 2*N   # Every cell divides, so the population size doubles
     output <- rbind(output,data.frame(Time=t,Cellcount=N)) # Add a row to the dataframe
     t<-t+1
  }
  return(output) # Return the dataframe
}

exp_growth=exponentialGrowth(P) # Run the simulation once, and store the result in "exp_growth"
 
## If you want to plot with base R, do this: plot(x=exp_growth$Time,y=exp_growth$Cellcount,type="l")

## But I like ggplot:
plot_norm <- exp_growth %>% 
  ggplot(aes(x=Time,y=Cellcount)) +
  geom_line(col="navy",size=1.5) +
  ggtitle("Exponential growth")

plot_ylog <- exp_growth %>% 
  ggplot(aes(x=Time,y=log(Cellcount))) +
  geom_line(col="navy",size=1.5) +
  ggtitle("Exponential growth")

plot_norm + plot_ylog
```

# Step 2) Including mutations

Extend the loop provided in step 1, to include mutations.

## Assumptions

* Mutations are neutral (mutants grow at the same rate at the wildtype)
* Mutations occur at every division at rate genomeLength * mutationRate (in both offspring)
* No more than 1 mutation occurs per division, and there are no secondary mutations (i.e. mutants don't mutate)

---

# Step 3) Only tracking RIF-resistant mutants, not ALL mutants

Extend your model to include ONLY mutations which infer rifampicin resistance. 

## Assumptions
* The total number of possible mutations is genomeLength * 3 (one for every substitution)
* The total number of substitutions that infer resistance is given by targetSize (either from Lindsey et al., or your own data)

---

# Step 4) Final code: repeating the experiment N times 

Now, let's repeat the experiment 132 times, like we did in the lab. 
